\section{Notations, basic notions}

In this first section we want to define the elementary notions that are used
throughout the whole book. We use the usual notions
\[ \mathbb{N} = \{ 0, 1, 2, \ldots \}\ \mbox{for the natural numbers} \]
\[ \mathbb{Z} = \{ \ldots, -2, -1, 0, 1, 2, \ldots \}\ \mbox{for the integer
numbers} \]
\[ \mathbb{Q} = \{ \frac{a}{b} \mid a,b \in \mathbb{Z}, b \neq 0 \}\ \mbox{for
the rational numbers } \]

For the set operations we use $\cup$ for the union and $\cap$ for the
intersection. Also $A \subset B$, $a \in A$, $a \not\in A$, $\bar{A}$, $A - B$,
$A \times B$ and $\emptyset$ have their usual meaning. For the power set of a
set $A$ we write $2^A$ or $Pot(A)$. $Card(A)$ denotes the cardinality of $A$.
Logical implication is denoted by $\Rightarrow$.

{\bf Mappings} are denoted as $f : A \rightarrow B$, in that case $f$ is a
total mapping. We write $Q(f) = A, Z(f) = B$. Here $Q$ stands for ''Quelle'' (source)
and $Z$ for ''Ziel'' (target).

If $f: A \rightarrow B, g : B \rightarrow C$ are mappings, then $f \circ g : A
\rightarrow C$ is the composed mapping that one gets by applying $f$ first and
then $g$, i.e. $(f \circ g)(a) = g(f(a))$. If $f:A \rightarrow B$ and $C
\subset A$, then $f(C) = \{ f(c) \mid c \in C \}$.

A subset $R \subset A \times B$ is called a {\bf relation} between $A$ and $B$.
$R_f \subset A \times B$ with $R_f = \{ (a,b) \mid f(a) = b \}$ is the relation
{\bf induced by} mapping f or the {\bf graph} of $f$.

Let $f : A \rightarrow B$ be a mapping, $A_1 \subset A$ and $g : A_1 \rightarrow
B$ a mapping. $f$ is called the {\bf continuation} of $g$ if $f(a_1) = g(a_1),
a_1 \in A_1$. In this case we also write $f \mid _{A_1} = g$ (in words: $f$
restricted to $A_1$).

A {\bf semi-group} consists of a set $M$ an an associative operation on that
set, usually denoted as a multiplication. If a semi-group is commutative, we
also use ''$+$'' instead of ''$\cdot$''.

A semi-group is a {\bf monoid} if $M$ contains a neutral element. We often
denote it with $1_M$ or shortly $1$. In the commutative case we often write $0$
instead of $1$.

For $A,B \subset M$ we denote by $A \cdot B = \{ a b \mid a \in A, b \in B \}$
the {\bf complex product} of $A$ and $B$.

$A \subset M$ is a {\bf submonoid} of $M$ if the follwoing holds: $1_M \in A$
and $A$ is closed under the operation of $M$.

For an $A$, the set $A^*$ defined as follows, is the smallest submonoid of $M$
which contains $A$. More specific,
\[
		A^* = \bigcap_{M' \in M(A)} M'	
\]
where $M(A) = \{ M' \subset M \mid M'$ is a submonoid of $M, A \subset M' \}$.

It is easy to see that
\[
		A^* = \bigcup_{n \geq 0} A^n\ \mbox{with}\ A^0 = \{1\}\ \mbox{and}\ A^{n+1} =
		A^n \cdot A
\]

In the same sense the notion $A^+ = A^* - \{1\}$ is defined for semi-groups. $A$
is called the {\bf generation system} of $A^*$ and $A^+$ resp.

A special meaning for us is assigned to the set of ''words'' (string) over a
fixed alphabet $A$. We understand as words the finite sequences of elements from
the alphabet $A$ as for example $(a,b,d,a,c)$ for alphabet $A = \{ a,b,c,d \}$.

We define
\[
WORD(A) := \{\epsilon\} \cup A \cup (A \times A) \cup (A \times A \times A) \cup
\ldots
\]
as the set of words (strings) over $A$. The symbol $\epsilon$ denotes the {\bf
empty word} over $A$, that is $A^0 = \{\epsilon\}$.

If $w, v \in WORD(A)$ then $w \cdot v$ is the word over $A$ which you get by
concatenating $w$ and $v$, more formally:
\[
\mbox{If}\ w = (a_1,\ldots,a_k), v = (a_{k+1},\ldots,a_n)\ \mbox{then}\ w \cdot
v = (a_1, \ldots, a_n)
\]

With this operation $WORD(A)$ becomes a monoid which is usually also denoted
with $A^*$. This is slightly inconsistent because for the first definition of
the $*$-operator it holds $(A^*)^* = A^*$ but for the second usage of the
$*$-operator it holds $(A*)^* \neq A^*$.

The following example should clarify that: Let $A = \{a,b,c\}$ and let $(a,b,a)$
and $(b,a) \in A^*$.
\[(a,b,a)\cdot(b,a) = (a,b,a,b,a) \in A^*,\ \mbox{but}\]
\[((a,b,a),(b,a)) \in (A^*)^*\ \mbox{but}\ \notin A^*\].


Instead of $(a)$ we write just $a$. In this sense it holds $A \subset A^*$. This
also holds in the sense of the first definition of $A^*$.

If $w = (w_1,\ldots,w_n)$ we denote with $|a| = n$ the {\bf length} of $w$.
Obviously it holds: $|w \cdot v| = |w| + |v|$ and $|\epsilon| = 0$.

The {\bf mirror word} $w^R$ of a word $w = (w_1,\ldots,w_n)$ is the word
$(w_n,\ldots,w_1)$. $\epsilon^R = \epsilon$. It holds: $(w \cdot v)^R =
v^R \cdot w^R$.

In $A^*$ the reduction rules hold, i.e.
\begin{enumerate}
  \item $w \cdot x = w \cdot y \Rightarrow x = y$
  \item $x \cdot w = y \cdot w \Rightarrow x = y$
\end{enumerate}

We define {\bf left} and {\bf right quotient}:
\[ B^{-1} \cdot A = \{ v\ |\ \exists u \in B, w \in A : u \cdot v = w \} \]
and $A \cdot B^{-1}$ in the same way.

Because of the reduction rules it holds:
\[ \{w\}^{-1} \cdot \{v\}\ \mbox{and}\ \{w\}^{-1} \cdot \{v\}\ \mbox{resp. are
either empty or contain a single element.} \]

If $\{w\}^{-1} \cdot \{v\}$ is not empty we call $w$ a {\bf prefix} of $v$. If
$\{w\} \cdot \{v\}^{-1} \not= \emptyset $, we call $v$ a {\bf suffix} of $w$. In
the future we will always write just $w$ instead of $\{w\}$ and also $w$ is
prefix of $v$ if $w^{-1} \cdot v \not= \emptyset$.

